---
title: "EDA"
format: html
editor: visual
---

## Introduction

The Diabetes Health Indicators Data set has 253,680 survey responses to the CDC's BRFSS2015 (Behavioral Risk Factor Surveillance System 2015). The primary response variable is "Diabetes_binary" which is in two classes: 0 for no diabetes and 1 for prediabetes/diabetes. Additionally, this data set has 21 variables.

For the purpose of this EDA and modeling, we will limit the analysis to explore 3 predictor variables in addition to our response variable of "Diabetes_binary". These include:

- **HighChol**: Adults who have been told they have high cholesterol. A class variable with 0 for no high cholesterol and 1 for high cholesterol.
- **BMI**: Body Mass Index ranging from 12 to 98 with an average of 28.4.
- **Age**: A 13-level age category where 1 = 18-24 all the way up to 13 = 80 and older and in between each level is in five year increments. The average age group is 55-59.

The first two were selected because of prior known relationships to them and diabetes. Age was selected because it would be interesting to know if it truly does have a relationship with the presence of diabetes.

Overall, the purpose of this EDA is to explore the chosen variables by summarizing them, looking at relationships, and evaluate different models to see their effectiveness in predicting prediabetes/diabetes or no diabetes. 

## Data

For the following section we are going to read in our data, convert the variables to factors if needed, and do any necessary data checks.

```{r warning=FALSE, message=FALSE}
## packages
library(tidyverse)

## read in the data
diabetes_data <- read.csv("diabetes_binary_health_indicators_BRFSS2015.csv")

## data type
str(diabetes_data)

## check for missing values
any(is.na(diabetes_data))
```

From this output we can see that we have a data frame with 253,680 observations of 22 variables. The output of "FALSE" tells us that there are no missing variables. We will now select only our variables of interest and convert them to factors when needed.

```{r}
diabetes_eda <- diabetes_data |>
  mutate(
    Diabetes = factor(Diabetes_binary, 
                             levels=c("0","1"),
                             labels=c("No","Yes")),
    HighChol = factor(HighChol,
                    levels=c("0","1"),
                    labels=c("No High Cholesterol","High Cholesterol")),
    Age = factor(Age,
               levels=c("1","2","3","4","5","6","7","8",
                            "9","10","11","12","13"),
               labels=c("18-24","25-29","30-34","35-39","40-44",
                        "45-49","50-54","55-59","60-64","65-69",
                        "70-74","75-79","80+"))) |>
  select(Diabetes, HighChol, BMI, Age)

str(diabetes_eda)
```

This is the data we will use for EDA.